---
output:
  github_document:
    html_preview: false
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
tmp <- tempfile()
dir.create(tmp)
```

# prt

<!-- badges: start -->
  [![Travis pkgdown build status](https://travis-ci.org/nbenn/prt.svg?branch=master)](https://travis-ci.org/nbenn/prt)
  [![Azure pipelines build status](https://img.shields.io/azure-devops/build/nbenn/prt/1)](https://dev.azure.com/nbenn/prt/_build/latest?definitionId=1&branchName=master)
  [![Azure pipelines test status](https://img.shields.io/azure-devops/tests/nbenn/prt/1?color=brightgreen&compact_message)](https://dev.azure.com/nbenn/prt/_build/latest?definitionId=1&branchName=master)
  [![Azure pipelines coverage status](https://img.shields.io/azure-devops/coverage/nbenn/prt/1)](https://dev.azure.com/nbenn/prt/_build/latest?definitionId=1&branchName=master)
<!-- badges: end -->

Building on `data.frame` serialization provided by [`fst`](https://www.fstpackage.org), `prt` offers an interface for working with partitioned `data.frame`s, saved as individual `fst` files.

## Installation

You can install the development version of [prt](https://nbenn.github.io/prt/) from GitHub by running

```{r gh-dev, eval = FALSE}
source("https://install-github.me/nbenn/prt")
```

Alternatively, if you have the `remotes` package available, the latest release is available by calling `install_github()` as

```{r gh-rel, eval = FALSE}
# install.packages("remotes")
remotes::install_github("nbenn/prt@*release")
```

## Short demo

Creating a `prt` object can be done either by calling `new_prt()` on a list of previously created `fst` files or by coercing a `data.frame` object to `prt` using `as_prt()`.

```{r create}
flights <- as_prt(nycflights13::flights, n_chunks = 2L, dir = tmp)
print(flights)
```

In case a `prt` object is created from a `data.frame`, the specified number of files is written to the directory of choice (a newly created directory within `tempdir()` by default).

```{r inspect}
list.files(tmp)
```

Subsetting and printing is closely modeled after `tibble` and behavior that deviates from that of `tibble` will most likely be considered a bug (please [report](https://github.com/nbenn/prt/issues/new)). Some design choices that do set a `prt` object apart from a `tibble` include the use of `data.table`s for any result of a subsetting operation and the complete disregard for `row.names`.

```{r cleanup, include = FALSE}
unlink(tmp, recursive = TRUE)
```
